Power and Sample Size in General Linear Mixed Models

### 2024-01-05 StatNet/LMM Power/statnet_lmm_power.qmd
LMM power presentation 99% done

```{r include=FALSE}
library(tidyverse)
library(simr)
```

@@

- Participants provided with AI-based cough monitor for 3 months.
  - Can distinguish between coughs and other noises.
- Cough frequency assessed by cough episodes per hour and cough seconds per hour.
- Compare the frequency of nighttime cough in study participants with CF versus healthy controls.
  - Controls will most likely be siblings of participants with CF.
  - This is the aim that we are powering on.
- Also interested in feasibility and the relationship between nighttime cough frequency and respiratory symptom scores.

@@

- Need to fill in $\rho$, $d$, and $\sigma^2$

@@

- Unfortunately there are very few previous studies on this.
  - Only found one study that reported variability of night time coughs in CF.
- According to Van Der Giessen et al.@vandergiessenNocturnalCoughChildren2009 the distributions of outcomes are skewed, so they report medians and IQR.
  - For the purposes of this power calculation we are assuming normality since we don't have access to the data.
- They report: 
  - 0.9 csec/hr (IQR 0.3-6.1) the first night and 0.96 csec/hr (IQR 0.1-3.4) the second night
  - Correlation between night 1 and night 2 was 0.29 for cough seconds and 0.35 for csec/hr.
- Assume that controls will not cough and pwCF will cough approximately 0.9-1 times per hour (same for csec/hr). Also assume that $\sigma = \frac{IQR}{1.35}$ and $\rho=0.3$

# Amalia's awesome workbook (demonstration)

@@

- Unfortunately not as easy to apply to this scenario since we don't have access to the prior data, but possible.
- First, set up a basic data matrix with 10 participants (one pwCF and one control from 5 families). Each person starts with 10 nights of data.
- Use the model:

@@

```{r}
# Set up X matrix with 10 nights of data for 5 families
nights <- 1:10
groups <- c("Control", "CF")
families <- as.character(1:5)
X <- expand.grid(time = nights, group = groups, family = families)
X$id <- paste0(X$group, X$family)
# Set simulation parameters
b <- c(0, 8, 0, 0) # Intercept and effect of night and group
V1 <- list(0.5, 2) # Random effect variance
# Create glmer object
glmer_obj <- makeGlmer(y ~ group * time + (1 | family / id),
  family = "poisson", fixef = b, VarCorr = V1, data = X
)
glmer_obj
```

```{r}
#| cache: true
powerSim(glmer_obj, nsim = 1000, test = "group", progress = F)
```

```{r}
#| cache: true
# Stable variables
nights <- 1:10
groups <- c("Control", "CF")
b <- c(0, 8, 0, 0) # Intercept and effect of night and group
V1 <- list(0.5, 2) # Random effect variance
# Alter sample size (number of families) and plot results
n <- lapply(2:100, function(f) {
  # Set up data
  families <- as.character(1:f)
  X <- expand.grid(time = nights, group = groups, family = families)
  X$id <- paste0(X$group, X$family)
  # Create glmer object
  glmer_obj <- makeGlmer(y ~ group * time + (1 | family / id),
    family = "poisson", fixef = b, VarCorr = V1, data = X
  )
  ps = powerSim(glmer_obj, nsim = 1000, test = "group", progress = F)
})

```

@@

- Do these assumptions seem reasonable given the limited prior knowledge?
- Does anyone have code they'd be willing to share for these simulations?
- Suggestions or software recommendations?

>>>>

```{r include=FALSE}
library(knitr)
library(tidyverse)
library(simr)
```

@@

- Participants provided with AI-based cough monitor for 3 months.
  - Can distinguish between coughs and other noises.
- Cough frequency assessed by cough episodes per hour and cough seconds per hour.
- First part of the study is feasibility.
- Also interested in the relationship between nighttime cough frequency and respiratory symptom scores.

- Compare the frequency of nighttime cough in study participants with CF versus healthy controls.
  - Controls will most likely be siblings of participants with CF.
  - This is the aim that we are powering on.
- We have repeated measures for each participant, plus an additional layer of clustering by family.
  - We expect some correlation between siblings.

## Power for a continuous outcome {.scrollable}

@@

- Need to fill in $\rho$, $d$, and $\sigma^2$ based on prior knowledge.

@@

- Unfortunately there are very few previous studies on this.
  - Only found one study that reported variability of night time coughs in CF.
- According to Van Der Giessen et al. @vandergiessenNocturnalCoughChildren2009 the distributions of outcomes are skewed, so they report medians and IQR.
  - For the purposes of this power calculation we are assuming normality since we don't have access to the data.

## Previous studies {.scrollable}

- They report: 
  - 0.9 csec/hr (IQR 0.3-6.1) the first night and 0.96 csec/hr (IQR 0.1-3.4) the second night
  - Correlation between night 1 and night 2 was 0.29 for cough seconds and 0.35 for csec/hr.

- We assume that controls will not cough @munyardHowMuchCoughing1996 and pwCF will cough approximately 0.9-1 times per hour (same for csec/hr).
- Correlation between successive nights is $\rho=0.3$.
- We use $\sigma = \frac{IQR}{1.35}$
  - This feels a little hand-wavy so please suggest alternatives.

# Amalia's awesome workbook

@@

- Unfortunately not as easy to apply to this scenario since we don't have access to the prior data, but theoretically possible.
- First, set up a basic data matrix with 10 participants (one pwCF and one control from 5 families). Each person starts with 10 nights of data.

## SIMR setup

- Using the model:

@@

## SIMR setup

```{r}
#| echo: TRUE
# Set up X matrix with 5 nights of data for 5 families
nights <- 1:5
groups <- c("Control", "CF")
families <- as.character(1:5)
X <- expand.grid(time = nights, group = groups, family = families)
X$id <- paste0(X$group, X$family)
```

## SIMR setup {.scrollable}

```{r}
kable(head(X,20))
```

## SIMR setup {.scrollable}

```{r}
#| echo: TRUE
# Set simulation parameters
b <- c(0, 1, 0, 0) # Intercept and effects of night and group
V1 <- list(0.5, 2) # Random effect variance
# Create glmer object
glmer_obj <- makeGlmer(y ~ group * time + (1 | family / id),
  family = "poisson", fixef = b, VarCorr = V1, data = X
)
glmer_obj
```

## SIMR results

```{r}
#| echo: TRUE
#| cache: true
powerSim(glmer_obj, nsim = 1000, test = "group", progress = F)
```

@@

- Do these assumptions seem reasonable given the limited prior knowledge?
- Is anyone familiar with `SIMR`? Would you be willing to share some code?
- Other general suggestions or software recommendations?
